---
title: "Podosenin-Proposal"
author: "Ursula Podosenin"
date: "2024-04-04"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Project Proposal

Dataset: Patient-level Clinical Drug Trial Data
https://www.kaggle.com/datasets/dillonmyrick/bells-palsy-clinical-trial

## Research Question?

Are gender and age predictors of recovery regardless of treatment?

## What are the cases, and how many are there?

Each case represents a patient. There are 494 cases

## Describe the method of data collection.

The data is housed on Kaggle and adapted from a clinical trial for Bell's Palsy: https://www.nejm.org/doi/full/10.1056/nejmoa072006#. The data collected studies the effect of two drugs, Prednisolone or Acyclovir in the early treatment of Bell’s Palsy 

## What type of study is this (observational/experiment)?

This is an experimental study with a medical intervention in the form of medication. 

## Data Source:
Found on Kaggle and adapted from a clinical trial for Bell's Palsy: https://www.nejm.org/doi/full/10.1056/nejmoa072006#. The data collected studies the effect of two drugs, Prednisolone or Acyclovir in the early treatment of Bell’s Palsy 

## What is the response variable, and what type is it (numerical/categorical)?

The response variable is the outcome, which is ordinal

## What is the explanatory variable, and what type is it (numerical/categorical)?

The explanatory variables are the gender and age of the participants in the control group, where gender is categorical and age is numerical.   

### Relevant summary statistics
```{r}

library(tidyverse)
library(ggplot2)
library(dplyr)
```

```{r}

# Reading the url file and creating a data frame
url= "https://raw.githubusercontent.com/ursulapodosenin/Projects/main/Bells%20Palsy%20Clinical%20Trial.csv"
BellsPalsyData<-as.data.frame(read.csv(url))
glimpse(BellsPalsyData)

# Creating a function that provides the value of the mode
Mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

# Calculating the total number of males and females in the study 
total_females<-sum(BellsPalsyData=="Female")
total_males<-sum(BellsPalsyData=="Male")
total_females
total_males

# Calculating the total number of participants in the control group and comparing it to the total number of participants
control_total<-sum(BellsPalsyData$Treatment.Group== "Placebo–Placebo")
control_total
noncontrol_total<-sum(BellsPalsyData$Treatment.Group!= "Placebo–Placebo")
noncontrol_total
control_ratio<-(control_total/noncontrol_total)
control_ratio

# Obtaining summary statistics regarding age, grouped by gender
summ_stats_by_sex<-
  BellsPalsyData|>
    group_by(Sex)|>
     summarise(mean_age= mean(Age),
              median_age= median(Age),
              mode_age= Mode(Age),
              min_age= min(Age),
              max_age= max(Age),
              sd_age= sd(Age) )
summ_stats_by_sex

# Creating an additional age group column that groups all participants into an age category 
BellsPalsyData['agegroup'] = case_when(BellsPalsyData$Age >= 10  & BellsPalsyData$Age <= 19 ~ '10-19',
                                             BellsPalsyData$Age >= 20  & BellsPalsyData$Age <= 29 ~ '20-29',
                                             BellsPalsyData$Age >= 30  & BellsPalsyData$Age <= 39 ~ '30-39',
                                             BellsPalsyData$Age >= 20  & BellsPalsyData$Age <= 29 ~ '20-29',
                                             BellsPalsyData$Age >= 30  & BellsPalsyData$Age <= 39 ~ '30-39',
                                             BellsPalsyData$Age >= 40  & BellsPalsyData$Age <= 49 ~ '40-49',
                                             BellsPalsyData$Age >= 50  & BellsPalsyData$Age <= 59 ~ '50-59',
                                             BellsPalsyData$Age >= 60  & BellsPalsyData$Age <= 69 ~ '60-69',
                                             BellsPalsyData$Age >= 70  & BellsPalsyData$Age <= 79 ~ '70-79',
                                             BellsPalsyData$Age >= 80  & BellsPalsyData$Age <= 89 ~ '80-89',
                                             BellsPalsyData$Age >= 90  & BellsPalsyData$Age <= 99 ~ '90-99')

# Creating a graph that looks at the distribution of age groups in the study 
ggplot(data=BellsPalsyData, aes(x=agegroup))+
  geom_bar(binwidth = 5, color="black", fill="blue")+
  theme_classic()+
  labs(title = "Frequency of Age Groups", x = "Age Group", y = "Frequency") 

# The House–Brackmann score is a score to grade the degree of nerve damage in a facial nerve palsy
# Looking at the most common time frame between the onset of symptoms and start of treatment
Mode_BP_Data<- Mode(BellsPalsyData$Time.between.onset.of.symptoms.and.start.of.treatment)
Mode_BP_Data

# Creating an additional column that looks the difference in recovery from baseline compared to 9 months of treatment
BellsPalsyData['difference']= BellsPalsyData$Baseline.Score.on.House.Brackmann.scale
-BellsPalsyData$X9.Month.Score.on.House.Brackmann.scale

# Creating a graphical representation of the difference between the ending score on the House Brackmann scale and the starting score
ggplot(data=BellsPalsyData, aes(x=difference))+
  geom_bar(binwidth = 10, color="black", fill="green")+
  theme_classic()+
  labs(title = "House Brackmann Score 9 Months After Initial Score", x = "House Brackmann Score", y = "Frequency") 


# Getting summary statistics by treatment group 
initial_summ_stats_by_treatment_group<-as.data.frame(
  BellsPalsyData|>
    group_by(Treatment.Group)|>
     summarise(mean_bs= mean(Baseline.Score.on.House.Brackmann.scale),
              median_bs= median(Baseline.Score.on.House.Brackmann.scale),
              mode_bs= Mode(Baseline.Score.on.House.Brackmann.scale),
              min_bs= min(Baseline.Score.on.House.Brackmann.scale),
              max_bs= max(Baseline.Score.on.House.Brackmann.scale),
              sd_bs= sd(Baseline.Score.on.House.Brackmann.scale) ))
initial_summ_stats_by_treatment_group
nine_summ_stats_by_treatment_group<-as.data.frame(
  BellsPalsyData|>
    group_by(Treatment.Group)|>
     summarise(mean_9s= mean(X9.Month.Score.on.House.Brackmann.scale),
              median_9s= median(X9.Month.Score.on.House.Brackmann.scale),
              mode_9s= Mode(X9.Month.Score.on.House.Brackmann.scale),
              min_9s= min(X9.Month.Score.on.House.Brackmann.scale),
              max_9s= max(X9.Month.Score.on.House.Brackmann.scale),
              sd_9s= sd(X9.Month.Score.on.House.Brackmann.scale) ))
nine_summ_stats_by_treatment_group

# Looking at the difference in means across treatment groups from the initial score to the 9 month score 
nine_summ_stats_by_treatment_group$mean_9s-initial_summ_stats_by_treatment_group$mean_bs

```



